A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship.}

### 2022-01-21 tex/esop22/examples.tex
[ minor ] changes made during meeting
[ rework ] usage elab above den sem, and 01ω

@@

@@

\begin{example}\label{thm:minus}
  We can make a subtraction function from primitive addition and negation on
  integers.
  Subtraction is covariant in its first argument and contravariant in its
  second argument.
  We give the definition in pseudocode, as we have not yet seen how to
  conveniently write terms (\cref{sec:usage-elaborator}).

@@

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We achieved some abbreviation by using pattern synonyms, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

To actually use \AgdaFunction{elab-sem} on terms, we take the associated
\AgdaFunction{semantics} and pass it the identity environment (an identity
\emph{renaming} in this case, because $\V$ is a family of variables).
We use \AgdaFunction{elab-unique}, which further
checks statically that exactly one derivation is returned.
The candidate usage vector \AgdaBound{R} will be empty for closed terms, and
otherwise we have to supply the intended usage annotations.

The example below is the term from \cref{thm:minus}.
We have instantiated the usage elaborator so that \AgdaField{0\#$^{-1}$},
\AgdaField{+$^{-1}$}, and \AgdaField{*$^{-1}$} are always singletons (i.e.,
deterministic), and \AgdaField{1\#$^{-1}$} is a singleton whenever it can be
(and empty for $\downarrow\downarrow$ and $\wn\wn$).
The notation \AgdaFunction{\#}\AgdaSpace{}\AgdaBound{i} turns statically known
natural number \AgdaBound{i} into the corresponding variable pointer, with
\AgdaNumber{0} being the leftmost.
In other words, we can write variables using de Bruijn levels.
The definition \AgdaFunction{\#0} is a workaround for an Agda bug.

\ExecuteMetaData[\Monotonicitytex]{minus}

>>>>

TODO: introductory paragraph

@@

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We can achieve some abbreviation by using pattern synonyms to wrap around
\AgdaInductiveConstructor{`con} expressions, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

To actually use \AgdaFunction{elab-sem} on terms, we take the associated
\AgdaFunction{semantics} and pass it the identity environment (an identity
\emph{renaming} in this case, because $\V$ is a family of variables).
We use \AgdaFunction{elab-unique}, which further
checks statically that exactly one derivation is returned.
The candidate usage vector \AgdaBound{R} will be \AgdaFunction{[]} for closed
terms, and otherwise we have to supply the intended usage annotations.

The example below is of a term needed to show that, in the
$\{\gr0, \gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
$\oc\gr\omega$ forms a comonad.
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and empty on
$\gr1$;
\AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and $\gr\omega$, and empty on
$\gr0$;
\AgdaField{+$^{-1}$} gives $\gr0 \mapsto [(\gr0,\gr0)]$,
$\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and
\AgdaField{*$^{-1}$} gives $(\gr\omega, \gr0) \mapsto [\gr0]$,
$(\gr\omega, \gr1) \mapsto []$, and
$(\gr\omega, \gr\omega) \mapsto [\gr\omega]$
(omitting $(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity).
Note that we do not consider splitting $\gr\omega$ up as, say,
$\gr1 + \gr\omega$, because this splitting would introduce more
non-determinism but not allow any more terms to be typed.
As such, the only non-determinism comes when we have variables annotated
$\gr1$ and need to do an additive split, like when we apply the
\AgdaInductiveConstructor{!E} rule below.
At this point, the variable can become either $\gr0$-annotated in the left
subterm and $\gr1$-annotated on the right, or vice-versa.
We will find that, because the left subterm wants to use that variable, the
former choice will be rejected.
The function \AgdaFunction{var\#} is a convenience for converting statically
known natural numbers, representing de Bruijn \emph{levels}, into variable
terms.

\ExecuteMetaData[\PaperExamplestex]{cojoin}

@@

\begin{example}\label{thm:minus}
  We can make a subtraction function from primitive addition and negation on
  integers.
  Subtraction is covariant in its first argument and contravariant in its
  second argument.
  We give the definition in pseudocode, though it is also amenable to the
  usage elaborator of \cref{sec:usage-elaborator}, suitably instantiated.

@@
